---
title: "Homework 2"
author: "Yige Bian (yb2591)"
date: 2023-10-01
output: github_document
---

```{r setup, include = FALSE}
Sys.setenv(LANG = "en_US")
library(tidyverse)
library(readxl)
```
# Problem 1
```{r tidy ppols_month, warning=FALSE}
# Import data "pol-month" and define types of columns. 
# Warning: There are several '2' in column 'prez_gop' which should be treated the same as '1'.  
# Break up the variable 'mon', replace month number with month name, create 'president' variable, remove 'prez_dem' and 'prez_gop' and 'day'

pols_month = 
  read_csv("./data/data_FiveThirtyEight/pols-month.csv", col_types = "Dliiiliii") |>
  janitor::clean_names() |>
  mutate(prez_gop = replace(prez_gop, is.na(prez_gop), TRUE)) |>
  separate(mon, into = c("year","month","day"), sep = "-") |>
  mutate(
    month = 
      case_match(
        month,
        "01" ~ "January",
        "02" ~ "February",
        "03" ~ "March",
        "04" ~ "April",
        "05" ~ "May",
        "06" ~ "June",
        "07" ~ "July",
        "08" ~ "August",
        "09" ~ "September",
        "10" ~ "October",
        "11" ~ "November",
        "12" ~ "December"),
    president = case_match(
        prez_gop,
        TRUE ~ "gop",
        FALSE ~ "dem")) |>
  select(year, month, everything(), -day, -starts_with("prez")) 
```
```{r tidy snp}
# Import and clean data snp 
snp = 
  read_csv("./data/data_FiveThirtyEight/snp.csv", col_types = cols(date = col_date("%m/%d/%y"))) |>
  janitor::clean_names() |>
  separate(date, into = c("year","month","day"), sep = "-") |>
  mutate(
    year = 
      as.character(ifelse(as.numeric(year) >= 2050, as.numeric(year)-100, as.numeric(year)))) |>
  arrange(year, month) |>
  select(year, month, everything()) |>
  mutate(
    month = 
      case_match(
        month,
        "01" ~ "January",
        "02" ~ "February",
        "03" ~ "March",
        "04" ~ "April",
        "05" ~ "May",
        "06" ~ "June",
        "07" ~ "July",
        "08" ~ "August",
        "09" ~ "September",
        "10" ~ "October",
        "11" ~ "November",
        "12" ~ "December"))
```

```{r tidy unemployment}
# Import data unemployment and swith from wide to long format. 
unemployment =
  read_csv("./data/data_FiveThirtyEight/unemployment.csv", col_types = "cd") |>
  janitor::clean_names() |>
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "per_of_unemp") |>
  mutate(
    month = 
      case_match(
        month,
        "jan" ~ "January",
        "feb" ~ "February",
        "mar" ~ "March",
        "apr" ~ "April",
        "may" ~ "May",
        "jun" ~ "June",
        "jul" ~ "July",
        "aug" ~ "August",
        "sep" ~ "September",
        "oct" ~ "October",
        "nov" ~ "November",
        "dec" ~ "December"))
```

```{r merge pols_month, snp and unemployment}
merge_data = 
  left_join(pols_month, snp, by = c("year","month")) |>
  left_join(unemployment, by = c("year","month"))
```

After tidying three datasets, the dataset `pols-month` contains 822 observations of 9 variables related to the number of national politicians who are democratic or republican from 1947 to 2015; The dataset `snp` contains 787 observations of 2 variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole from 1950 to 2015; The dataset `unemployment` contains 68 observations of 13 variables which describes the percentage of unemployment in each month from 1948 to 2015.
The resulting dataset merged by these three datasets has `r nrow(merge_data)` rows and `r ncol(merge_data)` columns, there are 822 observations and 12 varaibles, the range of years is `r range(merge_data$year)`. We observed that the range of years are different in three datasets, so when we merge them with left_join based on the consequence of `pols_month`, `snp` and `unemployment`, the values of variables in `snp` would be NA in 1947-1949, and values of variables in `unemployment` would be NA in 1948 because these two datasets don't have data for these years. From 1950 to 2015, the average unemployment rate was `r mean(merge_data$per_of_unemp[as.numeric(merge_data$year)>=1975 & merge_data$president=='dem'])` when a democrat was president. The average unemployment rate over the same time period in which a republican was president was `r mean(merge_data$per_of_unemp[as.numeric(merge_data$year)>=1975 & merge_data$president=='gop'])`.

# Problem 2
```{r Import and merge trash wheel data sets}
Mr_Trash_Wheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N549") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = 
      weight_tons*500/30,
    which_Trash_Wheel = "Mr")

Prof_Trash_Wheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M96") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = 
      weight_tons*500/30,
    year = as.character(year),
    which_Trash_Wheel = "Prof")

Gwyn_Trash_Wheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range="A2:K27") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = 
      weight_tons*500/30,
    year = as.character(year),
    which_Trash_Wheel = "Gwyn")

Mr_Prof_Gwyn=bind_rows(Mr_Trash_Wheel,Prof_Trash_Wheel,Gwyn_Trash_Wheel)
```

The data set `Mr. Trash Wheel` has 547 observations and 15 variables, the data set `Professor Trash Wheel` has 94 observations and 14 variables, the data set `Gwynnda Trash Wheel` has 25 observations and 12 variables. All these three data sets includes information on the dumpter number, date of collection, amount of total litter, amount of each litter type, and estimated homes powered.
The merged data of these three data sets has 666 observations and 16 variables. It is obvious that the observations of the merged data is the sum of observations of three data sets. The merged data contains variable `dumpster` which indicates the no. of dumpster and variables `year`, `month` and `date` which indicate the date, then `weight_tons` and viriables of different litter type show the amount of total litter and amount of each litter type. The total weight of trash collected by Professor Trash Wheel is: `r sum(Mr_Prof_Gwyn$weight_tons[Mr_Prof_Gwyn$which_Trash_Wheel=="Prof"])`, the total number of cigarette butts collected by Gwynnda in July of 2021 is: `r sum(Mr_Prof_Gwyn$cigarette_butts[Mr_Prof_Gwyn$which_Trash_Wheel=="Gwyn" & Mr_Prof_Gwyn$year=="2021" & Mr_Prof_Gwyn$month=="July"])`.

# Problem 3
```{r MCI_baseline, warning=FALSE}
MCI_baseline = 
  read_csv("./data/data_mci/MCI_baseline.csv", skip = 1, col_types = "cdlild") |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex,
        FALSE ~ "Female",
        TRUE ~ "Male"),
    apoe4=
      case_match(
        apoe4,
        TRUE ~ "carrier",
        FALSE ~ "non-carrier"))|>
  filter(is.na(age_at_onset) | age_at_onset - current_age > 0)
```

The original data set `MCI_baseline` has 483 observations and 6 variables which describes the Age at the study baseline, sex, years of education, apoe4 carry status and Age at the onset of MCI. Thorugh the data clean and tidy process, values of `sex` and `apoe4` were recoded to Female & Male and carrier & non-carrier based on their logical values. Then, if the `age_at_onset` is less than or equal to `current_age`, the correspond participants are not meet the stated inclusion criteria because they have MCI at baseline, and these observations (4 observations) were removed.
After such process, there are 479 observations left which means there are 479 participants were recruited who meet the stated inclusion criteria, and `r sum(!is.na(MCI_baseline$age_at_onset))` of them develop MCI during the follow-up period. The average baseline age is `r mean(MCI_baseline$current_age)`, `r sum(MCI_baseline$apoe4 == "carrier" & MCI_baseline$sex == "Female")/479`and of women in the study are APOE4 carriers.

```{r mci_amyloid, warning=FALSE}
mci_amyloid = 
  read_csv("./data/data_mci/mci_amyloid.csv", skip = 1, col_types = "cddddd") |>
  janitor::clean_names() |>
  mutate(id = study_id) |>
  select(id, everything(), -study_id)
```
The data set `mci_amyloid` has 487 observations and 6 variables, participants with id `setdiff(MCI_baseline$id, mci_amyloid$study_id)` appear in `MCI_baseline` but not in `mci_amyloid`, while participants with id `setdiff(mci_amyloid$study_id, MCI_baseline$id)` appear in `mci_amyloid` but not in `MCI_baseline`. This happened because 1. some participants in `MCI_baseline` who didn't meet the stated inclusion criteria were removed before; 2. some participants may lost contact during the follow-up period.

```{r merge two MCI datasets}
merged_MCI = inner_join(MCI_baseline, mci_amyloid, by = "id")
write.csv(merged_MCI, file = "./data/data_mci/merged_MCI.csv")
```

The merged data has 471 observations and 11 variables. The average age at study baseline in this new merged data is `r mean(pull(merged_MCI, current_age))`, the average amyloid $\beta$ 42/40 is `r mean(merged_MCI$baseline[!is.na(merged_MCI$baseline)])` (one baseline value of NA is excluded).




